---
title: "Cluster Analysis — stream"
author: "Hanyu Wang"
output: pdf_document
---

```{r, include=FALSE}
# install.packages(c("cluster", "factoextra", "fpc", "ggplot2", "gridExtra", "MASS", "NbClust", "readxl", "rlang"))

library(readxl)
library(cluster)
library(factoextra)
library(fpc)
library(ggplot2)
library(gridExtra)
library(MASS)
library(NbClust)
library(readxl)
```

## Introduction

The file `stream.xls` contains data on the prevalence of 11 species of
microcrustacea in seven streams in Alaska. Five measurements were made at
each site. The species are

  `sp1` Nitocra hibernica
  
  `sp2` Atheyella illinoisensis
  
  `sp3` Atheyella idahoenis
  
  `sp4` Bryocamptus hiemalis
  
  `sp5` Bryocamptus zschokkei
  
  `sp6` Acanthocyclops vernalis
  
  `sp7` Alona guttata
  
  `sp8` Graptoleberis
  
  `sp9` Chydorus
  
  `sp10` macrothricidae
  
  `sp11` Maraenobiotus insegnipes

Stream age, the Pfankuch index (a measure of stability – higher values are lower
stability), Temperature in degrees C, turbitidy, conductivity, and alkalinity were
also measured for each stream – these values are also included.

Data is taken from Peter Shaw’s Multivariate Statistics for the Environmental
Sciences (2003), p. 24

```{r, include=FALSE}
stream <- read_excel("stream.xls")
 
colnames(stream) <- c("stream", "rep",
                      "sp1","sp2","sp3","sp4","sp5","sp6",
                      "sp7","sp8","sp9","sp10","sp11",
                      "age","cwd","temp","turbidity","conductivity","alkalinity")
 
# Average the 5 replicates per stream
stream_mean <- aggregate(stream[, 3:19],
                         by = list(stream = stream$stream),
                         FUN = mean)
rownames(stream_mean) <- stream_mean$stream
stream_mean <- stream_mean[, -1]
 
sp_data  <- stream_mean[, 1:11]
env_data <- stream_mean[, 12:17]
```

## 1. DISTANCE METRICS AND DATA TRANSFORMATION

```{r}
sp_log <- log(sp_data + 1)
env_scaled  <- as.data.frame(scale(env_data))

combined    <- cbind(sp_log, env_scaled)
```

The species data (sp1–sp11) consist of count data with many zeros and a heavily right-skewed distribution. A log(x+1) transformation was applied prior to clustering to compress extreme values and produce a more symmetric distribution. The "+1" offset avoids undefined values when counts are zero. After log transformation, the species variables are on a comparable scale to one another, so no further standardization of the species data is needed.

The environmental variables (age, temperature, turbidity, conductivity, etc.) are measured on very different scales — for example, stream age ranges from 18 to 1,377 years while temperature ranges from 2.8 to 6.4°C. If left unstandardized, high-magnitude variables such as stream age would dominate the distance calculations regardless of their ecological relevance. Therefore, the environmental variables were standardized (z-scored) separately before being combined with the log-transformed species data.

Both Euclidean and Manhattan distances are appropriate for this combined continuous dataset. Euclidean distance is the standard choice; Manhattan distance is less sensitive to outliers and serves as a useful comparison. No binary or categorical distance metrics are needed since all variables are continuous after transformation.

## 2. HIERARCHICAL CLUSTER ANALYSIS

### Distance matrices

```{r}
#Get distance matrix
dist_euc <- dist(combined, method = "euclidean")   # Euclidean
dist_man <- dist(combined, method = "manhattan")   # Manhattan
```

### Cluster objects

```{r}
#Perform cluster analysis
euc_war <- hclust(dist_euc, method = "ward.D2")  # Euclidean + Ward
euc_com <- hclust(dist_euc, method = "complete") # Euclidean + Complete
euc_sin <- hclust(dist_euc, method = "single")   # Euclidean + Single

man_war <- hclust(dist_man, method = "ward.D2")  # Manhattan + Ward
man_com <- hclust(dist_man, method = "complete") # Manhattan + Complete
man_sin <- hclust(dist_man, method = "single")   # Manhattan + Single
```

### Plot dendrograms

```{r}
par(mfrow = c(2, 3), mar = c(4, 4, 3, 1))

plot(euc_war, xlab = "", ylab = "Distance", main = "Euclidean + Ward")
rect.hclust(euc_war, k = 3)

plot(euc_com, xlab = "", ylab = "Distance", main = "Euclidean + Complete")
rect.hclust(euc_com, k = 3)

plot(euc_sin, xlab = "", ylab = "Distance", main = "Euclidean + Single")
rect.hclust(euc_sin, k = 3)

plot(man_war, xlab = "", ylab = "Distance", main = "Manhattan + Ward")
rect.hclust(man_war, k = 3)

plot(man_com, xlab = "", ylab = "Distance", main = "Manhattan + Complete")
rect.hclust(man_com, k = 3)

plot(man_sin, xlab = "", ylab = "Distance", main = "Manhattan + Single")
rect.hclust(man_sin, k = 3)
```

Five of the six methods consistently identify the same three groups at the k=3 level:

  *Group A*: berg n, tyndall, berg s, rush pt
  *Group B*: carolus
  *Group C*: stonefly, wolf pt

  The exception is `Euclidean + Single`, which shows pronounced chaining — carolus splits off immediately at a very low distance, and the remaining streams merge sequentially rather than forming distinct groups. This is a well-known artifact of single linkage and makes the result difficult to interpret meaningfully.

Among the remaining five methods, the key difference lies in which two groups merge first at the higher level. 

  `Euclidean + Ward`, `Euclidean + Complete`, and `Manhattan + Ward` place Group B and Group C as closer to each other, merging carolus with stonefly/wolf pt before joining Group A. 

  `Manhattan + Complete` and `Manhattan + Single` place Group A and Group C as closer, merging berg n/tyndall/berg s/rush pt with stonefly/wolf pt on one branch.  

  This difference may reflect the greater sensitivity of Manhattan distance to the many zero-count species: stonefly and wolf pt each have almost no species, which makes their absolute count differences from the Bryocamptus-dominated Group A streams particularly large under the Manhattan metric, pushing them away from Group A and toward carolus instead.

Setting aside the unstable single linkage results, the remaining four methods agree on both the three-cluster solution, suggesting this grouping is robust.


## 3. EVALUATE NUMBER OF CLUSTERS

```{r}
#start with plot of universally applicable statistics
source("https://raw.githubusercontent.com/jreuning/sds363_code/refs/heads/main/HClusEval3.R.txt")

#Call the function
hclus_eval(combined, dist_m = 'euclidean', clus_m = 'complete', 
           plot_op = T, print_num = 5)
```
Using Euclidean distance and complete linkage, four metrics were examined to determine the optimal number of clusters.

  `RSQ`: Increases sharply from k=1 to k=2, then levels off, suggesting an elbow at k=2.

  `RMSSTD`: Drops steeply from k=3 to k=4, then stabilizes, indicating an elbow at k=4.
  
  `SPRSQ`: Drops steeply from k=1 to k=2, then stabilizes, indicating an elbow at k=2.

  `CD`: Drops steeply from k=1 to k=2, then stabilizes, indicating an elbow at k=2.


```{r}
# Note: NbClust with index = "all" failed because 
#       several indices require a covariance matrix, 
#       which cannot be computed with only 7 observations.

# NbClust(combined, distance = "euclidean", 
#             min.nc=2, max.nc=6, method = "complete", 
#             index = "all", alphaBeale = 0.1)

# I used ChatGPT to determine the indices that 
# do not require a covariance matrix

indices <- c("kl", "ch", "hartigan", "cindex", "db", 
             "silhouette", "duda", "pseudot2", "beale", "ratkowsky",
             "ball", "ptbiserial", "mcclain", "gamma", "gplus", 
             "tau", "dunn", "sdindex", "sdbw")

results <- list()
for (idx in indices) {
  tryCatch({
    res <- NbClust(combined, distance = "euclidean",
                   min.nc = 2, max.nc = 5, method = "complete",
                   index = idx, alphaBeale = 0.1)
    results[[idx]] <- res$Best.nc[1]
  }, error = function(e) {
    message(paste("Skipped:", idx, "-", e$message))
  })
}

# Summary
do.call(rbind, results)
```




Among 19 valid indices: 9 indices favored k=5, 5 indices favored k=2, 3 indices favored k=4, and 2 indices favored k=3. 

However, given that only 7 streams are available, k=5 or k=4 would leave some clusters with a single observation, making interpretation difficult. 

Combined with the results in Q2, k=2 was ruled out because three of the six dendrograms showed that one group would contain only a single stream, making it an uninterpretable partition. 

Therefore, k=3 is selected as the final solution.


## 4. K-Means Clustering


```{r}
set.seed(321)

km1 <- kmeans(combined, centers = 3)
km1

#see which countries are in each cluster
for (i in 1:3){
  print(paste("Streams in Cluster ", i))
  print(rownames(combined)[km1$cluster == i])
  cat("\n")
}
```




```{r}
#kdata is just normalized input dataset
kdata <- as.matrix(combined)
n.lev <- 6  #set max value for number of clusters k

# Calculate the within groups sum of squared error (SSE) 
# for the number of cluster solutions selected by the user
wss <- rnorm(10)
while (prod(wss==sort(wss,decreasing=T))==0) {
  wss <- (nrow(kdata)-1)*sum(apply(kdata,2,var))
  for (i in 2:n.lev) wss[i] <- sum(kmeans(kdata, centers=i)$withinss)}

# Calculate the within groups SSE for 250 randomized data sets 
# (based on the original input data)
k.rand <- function(x){
  km.rand <- matrix(sample(x),dim(x)[1],dim(x)[2])
  rand.wss <- as.matrix(dim(x)[1]-1)*sum(apply(km.rand,2,var))
  for (i in 2:n.lev) rand.wss[i] <- sum(kmeans(km.rand, centers=i)$withinss)
  rand.wss <- as.matrix(rand.wss)
  return(rand.wss)
}

rand.mat <- matrix(0,n.lev,250)

k.1 <- function(x) { 
  for (i in 1:250) {
    r.mat <- as.matrix(suppressWarnings(k.rand(kdata)))
    rand.mat[,i] <- r.mat}
  return(rand.mat)
}

# Same function as above for data with < 3 column variables
k.2.rand <- function(x){
  rand.mat <- matrix(0,n.lev,250)
  km.rand <- matrix(sample(x),dim(x)[1],dim(x)[2])
  rand.wss <- as.matrix(dim(x)[1]-1)*sum(apply(km.rand,2,var))
  for (i in 2:n.lev) rand.wss[i] <- sum(kmeans(km.rand, centers=i)$withinss)
  rand.wss <- as.matrix(rand.wss)
  return(rand.wss)
}

k.2 <- function(x){
  for (i in 1:250) {
    r.1 <- k.2.rand(kdata)
    rand.mat[,i] <- r.1}
  return(rand.mat)
}

# Determine if the data table has > or < 3 variables and call function above
if (dim(kdata)[2] == 2) { rand.mat <- k.2(kdata) } else { rand.mat <- k.1(kdata) }

# Plot within groups SSE against all tested cluster solutions

# Plot 1: Log SSE
xrange <- range(1:n.lev)
yrange <- range(log(rand.mat),log(wss))
plot(xrange,yrange, type='n', 
     xlab='Cluster Solution', ylab='Log of Within Group SSE', 
     main='Cluster Solutions against Log of SSE')
for (i in 1:250) lines(log(rand.mat[,i]),type='l',col='red')
lines(log(wss), type="b", col='blue')
legend('topright',c('Actual Data', '250 Random Runs'), 
       col=c('blue', 'red'), lty=1)

# Plot 2: Normal SSE
yrange <- range(rand.mat,wss)
plot(xrange,yrange, type='n', 
     xlab="Cluster Solution", ylab="Within Groups SSE", 
     main="Cluster Solutions against SSE")
for (i in 1:250) lines(rand.mat[,i],type='l',col='red')
lines(1:n.lev, wss, type="b", col='blue')
legend('topright',c('Actual Data', '250 Random Runs'), 
       col=c('blue', 'red'), lty=1)

# Calculate the mean and standard deviation of difference between SSE of actual 
# and randomized data
r.sse <- matrix(0,dim(rand.mat)[1],dim(rand.mat)[2])
wss.1 <- as.matrix(wss)
for (i in 1:dim(r.sse)[2]) {
  r.temp <- abs(rand.mat[,i]-wss.1[,1])
  r.sse[,i] <- r.temp}
r.sse.m <- apply(r.sse,1,mean)
r.sse.sd <- apply(r.sse,1,sd)
r.sse.plus <- r.sse.m + r.sse.sd
r.sse.min <- r.sse.m - r.sse.sd

# Plot differeince between actual SSE mean SSE from 250 randomized datasets

# Plot 3: Log SSE - Random SSE
xrange <- range(1:n.lev)
if (min(r.sse.min) < 0){
   yrange <- range(log(r.sse.plus - min(r.sse.min)*1.05), 
                   log(r.sse.min - min(r.sse.min)*1.05))
} else {
   yrange <- range(log(r.sse.plus), log(r.sse.min))
}

plot(xrange,yrange, type='n',
     xlab='Cluster Solution', ylab='Log of SSE - Random SSE', 
     main='Cluster Solutions against (Log of SSE - Random SSE)')
lines(log(r.sse.m), type="b", col='blue')
lines(log(r.sse.plus), type='l', col='red')
lines(log(r.sse.min), type='l', col='red')
legend('topright',c('SSE - random SSE', 'SD of SSE-random SSE'), 
       col=c('blue', 'red'), lty=1)

# Plot 4: Normal SSE - Random SSE
xrange <- range(1:n.lev)
yrange <- range(r.sse.plus,r.sse.min)
plot(xrange,yrange, type='n',
     xlab='Cluster Solution', ylab='SSE - Random SSE', 
     main='Cluster Solutions against (SSE - Random SSE)')
lines(r.sse.m, type="b", col='blue')
lines(r.sse.plus, type='l', col='red')
lines(r.sse.min, type='l', col='red')
legend('topright',c('SSE - random SSE', 'SD of SSE-random SSE'), 
       col = c('blue', 'red'), lty = 1)

# Apply K-means with chosen cluster number
clust.level <- 3

set.seed(321)

fit <- kmeans(kdata, clust.level)
aggregate(kdata, by=list(fit$cluster), FUN=mean)
clust.out <- fit$cluster
kclust <- as.matrix(clust.out)
kclust.out <- cbind(kclust, combined)
write.table(kclust.out, file="kmeans_out.csv", sep=",")

# See which streams are in each cluster
for (i in 1:clust.level){
  print(paste("Streams in Cluster", i))
  print(rownames(kdata)[fit$cluster == i])
  cat("\n")
}
```

K-means clustering was applied with k=3 (set.seed(321) for reproducibility). The plots suggest elbows at approximately k=2 and k=4.

The three-cluster solution explained 58.2% of total variance (between_SS / total_SS). The resulting groups are:

  `Cluster 1` (berg n, berg s, rush pt, tyndall): characterized by high sp4 (Bryocamptus hiemalis) and sp5 (B. zschokkei) abundance, moderate sp3 and sp6, and average environmental conditions.

  `Cluster 2` (carolus): a single-stream cluster distinguished by high sp1 (Nitocra hibernica), sp2 (Atheyella illinoisensis), sp6, sp8, and sp10, along with the oldest stream age — suggesting a uniquely mature, species-rich stream.
  
  `Cluster 3` (stonefly, wolf pt): characterized by near-zero species counts except sp6 and sp11 (Maraenobiotus insegnipes), high turbidity, and the coldest, youngest streams.

These results are consistent with the hierarchical clustering results from Questions 2 and 3.




## 5. Discriminant Analysis and PCA

```{r, include=FALSE}
sp_raw  <- stream[, 3:13]
env_raw <- stream[, 14:19]

sp_raw_log <- log(sp_raw + 1)
env_raw_scaled  <- as.data.frame(scale(env_raw))

combined_raw <- as.data.frame(scale(cbind(sp_raw_log, env_raw_scaled)))
```

```{r, include=FALSE}
dist_euc_raw <- dist(combined_raw, method = "euclidean")   # Euclidean
euc_war_raw <- hclust(dist_euc_raw, method = "ward.D2")  # Euclidean + Ward

cuts_raw <- cutree(euc_war_raw, k = 3)
```

### PCA space

```{r}
clusplot(combined_raw, cuts_raw, color = TRUE, shade = TRUE, labels = 2, lines = 0,
         main = "Three Cluster Plot, Complete Linkage, First two PC", cex = .5)
```

The three-cluster solution is displayed in PCA space. The first two principal components explain 43.8% of the total variability. The three clusters are well-separated, with Cluster 2 (carolus) appearing as a distinct outlier along Component 1, and Clusters 1 and 3 occupying different regions along Component 2.


### Discriminant analysis space

```{r}
plotcluster(combined_raw, cuts_raw, main = "Three Cluster Solution in DA Space",
            xlab = "First Discriminant Function", 
            ylab = "Second Discriminant Function")
```

DA space visualization was attempted but produced degenerate results. With only 35 observations and 17 variables, the within-group covariance matrix is nearly singular, causing DA to yield extreme, uninterpretable values. The PCA plot is therefore used as the sole visualization.


## 6. Number of Groups

Based on all analyses, three groups appear to be present in the data.

The hierarchical clustering dendrograms (Question 2), hclus_eval metrics (Question 3), and k-means SSE plots (Question 4) all suggested elbows at k=2 and k=4. The NbClust results were mixed, with votes spread across k=2, 3, 4, and 5. However,  k=3 remained the most interpretable choice given only 7 streams.

To further evaluate, PCA plots were produced for k=2 and 4. The k=2 solution merged ecologically distinct streams into one large overlapping cluster, losing meaningful differentiation. The k=4 solution visually resembles k=3, simply splitting one existing cluster further without revealing new structure. The k=3 solution produces the clearest separation — three distinct, non-overlapping groups — and is therefore selected as the final solution.

```{r}
dist_euc_raw2 <- dist(combined_raw, method = "euclidean")   # Euclidean
euc_war_raw2 <- hclust(dist_euc_raw2, method = "ward.D2")  # Euclidean + Ward

cuts_raw2 <- cutree(euc_war_raw2, k = 2)

clusplot(combined_raw, cuts_raw2, color = TRUE, shade = TRUE, labels = 2, lines = 0,
         main = "Two Cluster Plot, Complete Linkage, First two PC", cex = .5)
```

```{r}
dist_euc_raw4 <- dist(combined_raw, method = "euclidean")   # Euclidean
euc_war_raw4 <- hclust(dist_euc_raw4, method = "ward.D2")  # Euclidean + Ward

cuts_raw4 <- cutree(euc_war_raw4, k = 4)

clusplot(combined_raw, cuts_raw4, color = TRUE, shade = TRUE, labels = 2, lines = 0,
         main = "Four Cluster Plot, Complete Linkage, First two PC", cex = .5)
```

## 7. Description of Final Groups

The three final groups are consistent across hierarchical clustering and k-means, and make good ecological sense.

  `Group A` (berg n, berg s, rush pt, tyndall) represents moderate-aged, stable streams with diverse Bryocamptus communities. These four streams share high abundances of sp4 and sp5, along with moderate sp3 and sp6, low turbidity, and near-average environmental conditions. The co-occurrence of multiple Bryocamptus species suggests these are relatively stable, well-established streams that support a moderately diverse microcrustacea assemblage.

  `Group B` (carolus) is a single-stream cluster that stands out as ecologically distinct. Carolus has by far the oldest stream age and hosts the richest species composition, including sp1, sp2, sp6, sp8, and sp10 — species absent from all other streams. This is consistent with the expectation that older, more mature streams accumulate greater biodiversity over time.

  `Group C` (stonefly, wolf pt) represents young, disturbed streams with sparse microcrustacea communities. These streams have the youngest ages, lowest temperatures, highest turbidity, and near-zero species counts across most taxa, with only sp6 and sp11 present. High turbidity and young age suggest these streams are geomorphically unstable, supporting only the most tolerant species.

Overall, the three groups reflect a clear ecological gradient from young, disturbed, species-poor streams (Group C) through moderately developed streams (Group A) to an old, stable, species-rich stream (Group B), which is ecologically interpretable and consistent with stream succession theory.






